---
title: "Accessing DataForEver data using the SFNRC R package"
latex_engine: pdflatex
mainfont: FreeMono
output:
  pdf_document: null
  html_document:
    df_print: paged
subtitle: ''
header-includes:
- \usepackage{pdflscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
- \usepackage[utf8]{inputenc}
vignette: |
  %\VignetteIndexEntry{DataForEver} %\VignetteEngine{knitr::rmarkdown} %\usepackage[UTF-8]{inputenc}
---

```{r setup, include=FALSE, echo=FALSE}
pkgs.used <- c("knitr", "rmarkdown", "rgdal", "sp", "deldir", "plyr", "reshape2", "ggplot2", "scales", "gridExtra")
pkgs.to.install <- pkgs.used[!pkgs.used %in% installed.packages()]
if (length(pkgs.to.install) > 0) {
  install.packages(pkgs.to.install)
}
if (!"SFNRC" %in% installed.packages()) {
  remotes::install_github("troyhill/SFNRC")
}

# then load the package:
library(knitr)
library(rmarkdown)
library(rgdal)
library(sp)
library(deldir)
library(plyr)
library(reshape2)
library(ggplot2)
library(scales)
library(gridExtra)
library(SFNRC)
library(ggplot2) 
library(gridExtra)

knitr::opts_chunk$set(echo = TRUE, comment=NA)
```

\vspace{12pt}
\vspace{12pt}

## 1. Introduction

The `SFNRC` R package provides access to the DataForEver hydrology and water quality databases directly from R. This vignette demonstrates the use of  DataForEver-related functions in the `SFNRC` R package. Because DataForEver is hosted on internal servers of the South Florida Natural Resources Center, it can only be used when directly connected to the SFNRC's network. The elements of the `SFNRC` R package described in this vignette are not presently available to non-NPS users.


\vspace{12pt}

The `SFNRC` R package is hosted on GitHub and can be installed from the R console using the remotes package, as shown below. The package only needs to be installed once.

```{r install_SFNRC, include=TRUE, echo=TRUE, eval=FALSE}
### install devtools R package and a couple others that we'll use
if (!"ggplot2" %in% installed.packages()) install.packages("ggplot2")
if (!"gridExtra" %in% installed.packages()) install.packages("gridExtra")

### install SFNRC
if (!"SFNRC" %in% installed.packages())  remotes::install_github("troyhill/SFNRC")

library(SFNRC)
```



\vspace{3mm}\hrule
\vspace{12pt}



\vspace{12pt}
\vspace{12pt}

```{r, echo = FALSE, include=FALSE}
# set some constants
todaysDate <- substr(as.character(Sys.time()), 1, 10)
core.area <- pot.m2 <- 0.00801185 # mesocosm surface area (m2)
top.vol   <- core.area * 0.05 * 1e6 # cm3 in core: top 5 cm only
pointSize <- 2 # for ggplot graphics
pd <- pd2 <- position_dodge(1.2)
pd3 <- position_dodge(0.8)
grayColor <- 0.55
fig2Col   <- "gray55"


```

\vspace{12pt}
\vspace{12pt}


## 2. Automated access to DataForEver
\vspace{12pt}

Users can search for stations in the DataForEver water quality or hydrology databases. The optional `pattern` argument can be used to narrow the scope of the output.

\vspace{12pt}

```{r stn_query, include=TRUE, echo=TRUE, eval=FALSE}
getStn_DFE(dbname = "waterquality") # all WQ stations
getStn_DFE(pattern = "S333", dbname = "waterquality") # stations matching some string 
getStn_DFE(pattern = "S333", dbname = "hydrology")
```

\vspace{12pt}

Users can also query the parameters available from the hydrology or water quality databases. Queries of the water quality database can provide more concise, station-level lists of available parameters. If stations are used in the query, the output reports the number of observations for each station-parameter combination.


\vspace{12pt}

```{r stn_query2, include=TRUE, echo=TRUE, eval=FALSE}
### all parameters in water quality database
getParams_DFE(dbname = "waterquality")
getParams_DFE(dbname = "hydrology")
### parameters available at specific stations (note syntax)
getParams_DFE(stn = "S333|S18C", dbname = "waterquality")

```



\vspace{12pt}
\vspace{12pt}

## 3. Accessing DataForEver data
\vspace{12pt}

Using `SFNRC` to access DataForEver data is very straightforward. Data from the water quality database and hydrology databases are both accessed using the function `getDFE()`:

\vspace{12pt}


```{r wq, include=TRUE, echo=TRUE, results = "hide"}
# identify desired stations and water quality parameters
stations   <- c("S333", "S12A") # this notation is also fine: "S333|S12A"
wqParams   <- c("PHOSPH|NITROGEN|AMMONI|SUSPENDED|TURBIDITY")
hydroParams <- c("flow|head_water")

```

```{r wq2, include=TRUE, echo=TRUE, results = "hide", eval=FALSE}
flowDat <- getDFE(stn = stations, dbname = "hydrology", params = hydroParams,
            startDate = "2018-01-01")
 
wqDat <- getDFE(stn = stations, dbname = "waterquality", params = wqParams,
             startDate = "2018-01-01")
unique(wqDat$stn)

```

\vspace{12pt}

The `data_shape` argument reshapes data from long to wide form to facilitate a broad range of analyses. The default form is `long` (one row per date-station-param) but this can also be set to `wide` (one row per date-station) or `really_wide` (one row per date). Importantly, if there are multiple observations for a single date, they are averaged if the `data_shape` argument is set to `wide` or `really_wide`.

\vspace{12pt}

```{r hydro_get_data21, include=TRUE, echo=TRUE, eval=FALSE}
### by default, data are in "long" format, with one row 
### for every date-station-parameter combination
hyd.long  <- getDFE(stn = stations, params = hydroParams, startDate = "2018-01-01")

### but data can also be retrieved in "wide" (one row per date-station) or 
### "really_wide" (one row per date) forms, enabling more rapid 
### comparison between stations and/or parameters
hyd.wide  <- getDFE(stn = stations, params = hydroParams, startDate = "2018-01-01",
                      data_shape = "wide")
hyd.vwide <- getDFE(stn = stations, params = hydroParams, startDate = "2018-01-01",
                      data_shape = "really_wide")

```


```{r hydro_show, include=TRUE, echo=TRUE, eval=FALSE}
head(hyd.long) # one row for every date-station-parameter combination
head(hyd.wide) # one row for every date-station combination
head(hyd.vwide) # one row for every date

```



\vspace{12pt}
\vspace{12pt}

Merging water quality and hydrology data is a routine task that is automated by `SFNRC`. Instead of navigating the two databases and then figuring out how to merge the two datasets, water quality data can be downloaded at the same time as flow/hydrology data by setting `addWaterQuality = TRUE` in a call to `getDFE(dbname = 'hydrology')`. To reduce the number of columns in the returned dataset, it is recommended that water quality parameters be specified manually in this workflow, because the returned dataset will have two columns per parameter (returning the MDL and the value).

\vspace{12pt}

```{r hydro_get_data2, include=TRUE, message = FALSE, warning = FALSE, eval = FALSE}

### simultaneously download hydrology and water quality data:
hyd.wq <- getDFE(stn = stations, 
                   params = hydroParams, # hydrology database parameters
                   startDate = "2000-01-01",
                   addWaterQuality = TRUE, 
                   addWaterQualityParams = wqParams) # water quality database parameters

```

```{r internalWorkaround, include=TRUE, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE}
### simultaneously download hydrology and water quality data:
dd.wq <- plyr::ddply(wqDat[(wqDat$stn %in% stations) & grepl(x = wqDat$param, pattern = wqParams), names(wqDat) %in% c("stn", "date", "parameter", "value")], .(stn, date, parameter),
                         summarise,
                         value = mean(value, na.rm = TRUE))
wq.temp <- stats::reshape(dd.wq, 
                              idvar = c("stn", "date"), timevar = "parameter", 
                              direction = "wide")
    names(wq.temp) <- gsub(x = names(wq.temp), pattern = "value.| |,", 
                           replacement = "")

hyd.wq <- plyr::join_all(list(hydDat[hydDat$stn %in% stations, ], wq.temp), 
                              by = c("stn", "date"))
  
```

\vspace{12pt}

At this point the data can be plotted or analyzed. For example, simple time series plots:

\vspace{12pt}
```{r hydro_show_merged, include=TRUE, echo=TRUE, warning=FALSE, message=FALSE, fig.width = 3, fig.height = 3}

ggplot(data = data.frame(hyd.wq), aes(y = PHOSPHATETOTALASP, x = date, shape = stn)) + 
  theme_classic() + geom_point(show.legend = FALSE) + geom_smooth() + 
  facet_grid(stn ~ .) + ylim(0, 0.1) + ylab("Total P (mg/L)") + xlab("")

```

\vspace{12pt}

\pagebreak

Relationships between hydrologic and water quality parameters can be explored: 

\vspace{12pt}

```{r plot_wqFlow, echo=TRUE, warning=FALSE, message=FALSE, fig.width = 3, fig.height = 4}
CQ <- ggplot(data = data.frame(hyd.wq[(hyd.wq$flow > 0) & !is.na(hyd.wq$flow), ]), 
  aes(y = log(PHOSPHATETOTALASP), x = log(flow), shape = stn)) + 
  theme_classic() + geom_point(show.legend = FALSE) + geom_smooth(method = "lm") + 
  facet_grid(stn ~ .) + ylab("log(TP) (mg/L)") + xlab("log(discharge) (cfs)")

C_HW <- ggplot(data = data.frame(hyd.wq), aes(y = PHOSPHATETOTALASP, x = head_water, 
  shape = stn)) + theme_classic() + geom_point(show.legend = FALSE) + geom_smooth() + 
  facet_grid(stn ~ .) + ylim(0, 0.1) + ylab("log(TP) (mg/L)") + 
  xlab("Head water stage (feet NGVD29)")

grid.arrange(CQ, C_HW, nrow = 2)

```

\pagebreak

And relationships between water quality parameters can be evaluated:

\vspace{12pt}

```{r plot_wq_bivariate, echo=TRUE, warning=FALSE, message=FALSE, fig.width = 6, fig.height = 4}
TP.NTU <- ggplot(data = data.frame(hyd.wq), aes(y = PHOSPHATETOTALASP, x = TURBIDITY, 
  shape = stn)) + theme_classic() + geom_point(show.legend = FALSE) + geom_smooth(method = "lm") + 
  facet_grid(stn ~ .) + ylab("TP (mg/L)") + xlab("Turbidity (NTU)") + ylim(0, 0.1)

TN.TP <- ggplot(data = data.frame(hyd.wq), aes(y = PHOSPHATETOTALASP, x = TOTALNITROGEN, 
  shape = stn))  + theme(legend.position="none") + theme_classic() + 
  geom_point(show.legend = FALSE) + geom_smooth(method = "lm") + facet_grid(stn ~ .) + 
  ylab("TP (mg/L)") + xlab("TN (mg/L)") + ylim(0, 0.1) + xlim(0, 3)

grid.arrange(TP.NTU, TN.TP, ncol = 2)

```




\vspace{12pt}
\vspace{12pt}

